Sample cardinalR

### 2025-12-17 extra/paper-cardinalR-code.Rmd
Updated branch projections

### Cone

To simulate a cone-shaped structure in arbitrary dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cone)"} else { "Figure \\@ref(fig:cone-proj)"})`), we define a function `gen_cone(n, p, h, ratio)`, which creates a high-dimensional cone with options for a sharp or blunted apex, allowing for a dense concentration of points near the tip.

This function generates $n$ points in $p\text{-}D$, where the last dimension, $X_p$, represents the height along the cone's axis, and the first $p-1$ dimensions define a shrinking hyperspherical cross-section toward the tip. Heights are sampled from a truncated exponential distribution, $X_p \sim \text{Exp}(\lambda = 2/h)$, truncated to the interval $[0, h]$, producing a higher density of points near the tip. At each height $X_p$, the radius of the cross-section increases linearly from base to tip according to $r = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}}) X_p / h$, where $r_{\text{min}} = \text{ratio} \in [0, 1]$ and $r_{\text{max}} = 1$.

For each point, a direction in the first $p-1$ dimensions is sampled uniformly on a $(p-1)$-dimensional hypersphere using generalized spherical coordinates. The radial coordinates are scaled by the height-dependent radius $r$, producing the conical taper. In three dimensions ($p = 3$), this results in a classical $3\text{-}D$ cone, while for $p > 3$, additional dimensions provide a smooth embedding into higher-dimensional space, preserving the conical structure.

Cone-shaped structures appear in particle dispersions, light beams, and tapering processes, where spread decreases along one axis. They are also used to benchmark clustering and dimensionality reduction methods [@hadsell2006].

```{r cone-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from the $3\\text{-}D$ \\texttt{cone} data. Points are concentrated near the tip along the height dimension, while the radius of the hyperspherical cross-section decreases linearly toward the apex. These projections show how the conical geometry is preserved.", fig.pos="H", fig.pos="!ht", fig.width=12, fig.height=4}
cone <- gen_cone(n = 1000, p = 4, h = 5, ratio = 0.5)

@@

centered_data <- center_data(cone)

## First projection
projection <- cbind(
  c(0.22417,-0.02943,-0.02870,0.08418),
  c(-0.02715,0.18019,0.05053,0.15254))

proj_obj1 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.6,
                                              axis_scaled = 3, 
                                              axis_pos_x = -0.4, 
                                              axis_pos_y = -0.4, 
                                              threshold = 0.053))

cone_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.5, 0.7), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

## Second projection
projection <- cbind(
    c(0.15591,0.17907,0.03198,0.04496),
    c(-0.15376,0.10435,0.15766,0.00547))

proj_obj2 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.3,
                                              axis_scaled = 2, 
                                              axis_pos_x = -0.2, 
                                              axis_pos_y = -0.2, 
                                              threshold = 0.02))

cone_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.25, 0.35), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

## Third projection
projection <- cbind(
    c(-0.13548,0.13832,0.05991,-0.13544),
    c(0.08479,-0.10719,-0.01572,-0.20123))

proj_obj3 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.6,
                                              axis_scaled = 3, 
                                              axis_pos_x = -0.6, 
                                              axis_pos_y = -0.6, 
                                              threshold = 0.04))

cone_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.7, 0.5), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

cone_proj1 + cone_proj2 + cone_proj3 +
  plot_layout(ncol = 3, guides = "collect") 
```

>>>>

# Application

This section demonstrates how the package can be used to generate complex high-dimensional datasets, evaluate dimension reduction (DR) and clustering methods. The example shows how diverse geometric structures can be simulated and analyzed to assess algorithmic behavior.

To illustrate how high-dimensional clustered data can be generated using `cardinalR`, we generate a dataset with five clusters in $4\text{-}D$, each representing distinct geometric characteristics: a *helical spiral* (elongated and twisted), a *hemisphere* (curved surface), a *uniform cube* (isotropic distribution), a *cone* (density gradient), and a *Gaussian* cluster (compact and spherical) (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:highd-data)"} else { "Figure \\@ref(fig:highd-proj)"})`). Each cluster has a unique number of points and scaling factor, representing variation in cluster size and spread across the $4\text{-}D$ space.

```{r highd-proj, eval=knitr::is_latex_output(), fig.pos="!ht", fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the five clusters data. The helical spiral cluster is represented in dark green, the hemisphere cluster in orange, the uniform cube-shaped cluster in purple, the blunted cone cluster in pink, and the Gaussian-shaped cluster in light green.", fig.alt="Three $2\\text{-}D$ projections of five $4\\text{-}D$ synthetic clusters: helical spiral (dark green), hemisphere (orange), uniform cube (purple), cone (pink), and Gaussian (light green), showing spatial separation and cluster geometry.", fig.width=12, fig.height=4, fig.pos="!ht"}

@@

positions <- geozoo::simplex(p=4)$points
positions <- positions * 0.3

five_clusts <- gen_multicluster(n = c(2250, 1500, 750, 1250, 1750), k = 5,
                       loc = positions,
                       scale = c(0.25, 0.35, 0.3, 1, 0.3),
                       shape = c("helicalspiral", "hemisphere", "unifcube", 
                                 "cone", "gaussian"),
                       rotation = NULL,
                       is_bkg = FALSE)

centered_data <- center_data(five_clusts[, -5])

## First projection
projection <- cbind(
  c(-0.59692,0.46414,0.42436,-0.10048),
  c(-0.54999,-0.67673,-0.03631,-0.01193))

proj_obj1 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.5,
                                              axis_scaled = 0.8, 
                                              axis_pos_x = -0.4, 
                                              axis_pos_y = -0.4, 
                                              threshold = 0.02))

proj_obj1[["cluster"]] <- as.character(five_clusts$cluster)

five_clusts_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.5, 0.5), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")

## Second projection
projection <- cbind(
  c(0.34673,-0.35774,0.66272,0.27298),
  c(0.24023,0.01170,-0.41886,0.72707))

proj_obj2 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.55,
                                              axis_scaled = 0.8, 
                                              axis_pos_x = -0.4, 
                                              axis_pos_y = -0.4, 
                                              threshold = 0.05))

proj_obj2[["cluster"]] <- as.character(five_clusts$cluster)

five_clusts_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.5, 0.67), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")

## Third projection
projection <- cbind(
  c(0.47388,-0.20984,-0.56560,0.41644),
  c(-0.51124,-0.60184,0.06463,0.36627))

proj_obj3 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.5,
                                              axis_scaled = 0.8, 
                                              axis_pos_x = -0.35, 
                                              axis_pos_y = -0.35, 
                                              threshold = 0))

proj_obj3[["cluster"]] <- as.character(five_clusts$cluster)

five_clusts_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.45, 0.55), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")

five_clusts_proj1 + five_clusts_proj2 + five_clusts_proj3 +
  plot_layout(ncol = 3) 
```

```
